---
title: "Epistatic Nested Effects Models - Inferring mixed epistatis from indirect measurements of knockout screens."
author: "Madeline Diekmann & Martin Pirkl"
date: "`r Sys.Date()`"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{epiNEM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
## Introduction

This package is an extension of the classic Nested Effects Models provided in package \emph{nem}. Nested Effects Models is a pathway reconstruction method, which takes into account effects of downstream genes. Those effects are observed for every knockout of a pathway gene, and the nested structure of observed effects can then be used to reconstruct the pathway structure.
However, classic Nested Effects Models do not account for double knockouts. In this package \emph{epiNEM}, one additional layer of complexity is added. For every two genes, acting on one gene together, the relationship is evaluated and added to the model as a logic gate. Genetic relationships are represented by the logics OR (no relationship), AND (functional overlap), NOT (masking or inhibiting) and XOR (mutual prevention from acting on gene C).

## Loading epiNEM

```{r, eval=FALSE}
install.packages("devtools", verbose = F, quiet = T)

library(devtools)

install_github("cbg-ethz/epiNEM", quiet = T)
```
```{r}
library(epiNEM)
```

## Simulations

We compare epiNEM to several network inference methods.

```{r}
library(bnem, quietly = T, verbose = F) # install_github("MartinFXP/B-NEM/package")

library(nem)

library(minet)

library(pcalg)

runs <- 100

noiselvls <- c(0.01, 0.025, 0.05, 0.1, 0.2, 0.3, 0.4, 0.5)

random <- list(FPrate = 0.1, FNrate = noiselvls,
               single = 4, double = 1, reporters = 100, replicates = 3)

spec <- sens  <- logics <- array(0, dim = c(2, runs, length(noiselvls)))

sens2 <- spec2 <- time <- array(0, dim = c(5, runs, length(noiselvls)))

do <- c("n", "p", "a")

do <- c("e", "b", do)

popSize <- 100

maxTime <- F

forcelogic <- T

epinemsearch <- "greedy"

nIterations <- 3

bnemsearch <- "genetic"

parallel <- NULL

logicgate <- matrix("", runs, length(noiselvls))

edgenr <- matrix(0, runs, length(noiselvls))
```
```{r, eval=FALSE}
for (i in 1:runs) {

    print(paste("run ", i, sep = ""))

    for (j in 1:length(noiselvls)) {

        print(paste("noiselvl ", j, sep = ""))
        
        topology <- CreateTopology(random$single, random$double, force = forcelogic)

        topology <- unlist(unique(topology), recursive = FALSE)

        extTopology <- ExtendTopology(topology$model, random$reporters)

        sortedData <- GenerateData(topology$model, extTopology, 
                                   random$FPrate, random$FNrate[j], random$replicates)

        logicgate[i, j] <- paste(topology$logics, collapse = "_")

        edgenr[i, j] <- sum(topology$origModel == 1)

        if ("e" %in% do) {
            print("epiNEM")

            start <- Sys.time()
            TriplModel <- epiNEM(filename = sortedData,
                      method = epinemsearch, nIterations = nIterations)
            time[1, i, j] <- difftime(Sys.time(), start, units = "secs")
            print(time[1, i, j])

            tp <- sum(topology$model == 1 & TriplModel$model == 1)
            tn <- sum(topology$model == 0 & TriplModel$model == 0)
            fp <- sum(topology$model == 0 & TriplModel$model == 1)
            fn <- sum(topology$model == 1 & TriplModel$model == 0)
            sens[1, i, j] <- tp/(tp+fn)
            spec[1, i, j] <- tn/(tn+fp)
            tp <- sum(topology$origModel == 1 & TriplModel$origModel == 1)
            tn <- sum(topology$origModel == 0 & TriplModel$origModel == 0)
            fp <- sum(topology$origModel == 0 & TriplModel$origModel == 1)
            fn <- sum(topology$origModel == 1 & TriplModel$origModel == 0)
            sens2[1, i, j] <- tp/(tp+fn)
            spec2[1, i, j] <- tn/(tn+fp)
            tp <- 0
            for (k in 1:length(topology$column)) {
                for (l in 1:length(TriplModel$column)) {
                    if (topology$column[k] == TriplModel$column[l]) {
                        if (topology$logics[k] %in% TriplModel$logics[l]) {
                            tp <- tp + 1
                        }
                    }
                }
            }
            logics[1, i, j] <- tp/(length(topology$logics) +
                                      length(TriplModel$logics) - tp)
            print(sens[1, i, j])
            print(spec[1, i, j])
            print(sens2[1, i, j])
            print(spec2[1, i, j])
            print(logics[1, i, j])    

        }

        if ("b" %in% do) {
            print("B-NEM")

            gtn <- epi2bg(topology)

            fc <- cbind(Ctrl_vs_S = -1, epi2bg(sortedData))*(-1)

            bnemnoise <- sample(1:nrow(fc), floor(nrow(fc)*random$FNrate[j]))
            
            fc[bnemnoise, 1] <- 0
            
            ers <- t(topology$model)*(-1)
            colnames(ers) <- paste("S_vs_S_",
                                 gsub("\\.", "_", colnames(ers)), sep = "")
            ers <- cbind(Ctrl_vs_S = 1, ers)
            ers <- ers[, order(colnames(ers))]

            CNOlist <- dummyCNOlist(stimuli = "S",
                               inhibitors = LETTERS[1:random$single],
                               maxStim = 1, maxInhibit = 2,
                               signals = LETTERS[1:random$single])

            parents <- unique(unlist(strsplit(colnames(sortedData)[grep("\\.",
                               colnames(sortedData))], "\\.")))

            nodes <- unique(colnames(sortedData)[-grep("\\.", colnames(sortedData))])

            child <- nodes[-which(nodes %in% parents)]

            sifMatrix <- NULL
            for (k in LETTERS[1:random$single]) {
               sifMatrix <- rbind(sifMatrix, c("S", "1", k))#, c("S", "-1", k))
                for (l in LETTERS[1:random$single]) {
                    if (k %in% l) { next() }
                    if (k %in% parents) {
                        sifMatrix <- rbind(sifMatrix, c(k, "1", l), c(k, "-1", l))
                    } else {
                        sifMatrix <- rbind(sifMatrix, c(k, "1", l))
                    }
                
            randfile <- paste("pkn_", as.numeric(Sys.time()), sep = "")
            write.table(sifMatrix, file = randfile, sep = "\t",
                        row.names = FALSE, col.names = FALSE, quote = FALSE)
            PKN <- readSIF(randfile)
            unlink(randfile)

            model <- preprocessing(CNOlist, PKN)

            initBstring <- absorption(rep(1, length(model$reacID)), model)

            if (maxTime) { maxTime2 <- time[1, i, j] } else { maxTime2 <- Inf }

            start <- Sys.time()
            bga <- bnem(search = bnemsearch,
                        fc=fc,
                        CNOlist=CNOlist,
                        model=model,
                        initBstring=initBstring,
                        draw = F,
                        verbose = F,
                        popSize = popSize,
                        maxTime = maxTime2,
                        parallel = parallel
                        )
            time[2, i, j] <- difftime(Sys.time(), start, units = "secs")
            print(time[2, i, j])

            ers2 <- computeFc(CNOlist, t(simulateStatesRecursive(CNOlist,
                                                                 model, bga$bString)))
            ers2 <- ers2[, unique(colnames(fc))]
            ers2 <- ers2[, order(colnames(ers2))]

            tp <- sum(ers == -1 & ers2 == -1)
            tn <- sum(ers == 0 & ers2 == 0)
            fn <- sum(ers == -1 & ers2 == 0)
            fp <- sum(ers == 0 & ers2 == -1)
            sens[2, i, j] <- tp/(tp+fn)
            spec[2, i, j] <- tn/(tn+fp)
            gtn2 <- abs(dnf2adj(gtn))
            if (length(grep("S", rownames(gtn2))) > 0) {
                gtn2 <- gtn2[-grep("S", rownames(gtn2)), -grep("S", colnames(gtn2))]
            }
            gtn2 <- gtn2[order(rownames(gtn2)), order(colnames(gtn2))]
            res <- abs(dnf2adj(bga$graph))
            if (length(grep("S", rownames(res))) > 0) {
                res <- as.matrix(res[-grep("S", rownames(res)),
                                                    -grep("S", colnames(res))])
            }
            if (dim(res)[1] == 1) {
                colnames(res) <- rownames(res) <- gsub(".*=", "", bga$graph)
            } else {
                res <- res[order(rownames(res)), order(colnames(res))]
            }
            if (nrow(res) < nrow(gtn2)) {
                res2 <- rbind(cbind(res, matrix(0, nrow(res), nrow(gtn2) - nrow(res))),
                              matrix(0, nrow(gtn2) - nrow(res), ncol(gtn2)))
                colnames(res2)[(ncol(res)+1):ncol(res2)] <-
                                     colnames(gtn2)[which(!(colnames(gtn2)
                                     %in% colnames(res)))]
                rownames(res2)[(nrow(res)+1):nrow(res2)] <-
                                     rownames(gtn2)[which(!(rownames(gtn2)
                                     %in% rownames(res)))]
                res2 <- res2[order(rownames(res2)), order(colnames(res2))]
                res <- res2
            }
            diag(gtn2) <- diag(res) <- 0
            tp <- sum(gtn2 == 1 & res == 1)
            tn <- sum(gtn2 == 0 & res == 0)
            fn <- sum(gtn2 == 1 & res == 0)
            fp <- sum(gtn2 == 0 & res == 1)
            sens2[2, i, j] <- tp/(tp+fn)
            spec2[2, i, j] <- tn/(tn+fp)
            tp <- sum(bga$graph %in% gtn)
            logics[2, i, j] <- tp/(length(gtn) + length(bga$graph) - tp)
            print(sens[2, i, j])
            print(spec[2, i, j])
            print(sens2[2, i, j])
            print(spec2[2, i, j])
            print(logics[2, i, j])

            print(bga$graph)
            print(gtn)

        }

        if (any(c("n", "p", "a") %in% do)) {

            reddata <- sortedData[, -grep("\\.", colnames(sortedData))]
            gtnadj <- topology$origModel
            gtnadj <- gtnadj[order(apply(gtnadj, 1, sum), decreasing = T),
                                        order(apply(gtnadj, 2, sum), decreasing = F)]
            gtnadj[lower.tri(gtnadj)] <- gtnadj[upper.tri(gtnadj)]
            gtnadj <- gtnadj[order(rownames(gtnadj)), order(colnames(gtnadj))]
            eadj <- topology$origModel
            eadj <- eadj[order(rownames(eadj)), order(colnames(eadj))]
            reddata2 <- matrix(0, nrow(reddata)*random$replicates,
                                         length(unique(colnames(reddata))))
            for (k in 1:length(unique(colnames(reddata)))) {
                reddata2[, k] <- as.vector(reddata[, which(colnames(reddata) %in%
                                                     unique(colnames(reddata))[k])])
            }
            colnames(reddata2) <- unique(colnames(reddata))

        }

        if ("n" %in% do) {
            print("NEM")

            start <- Sys.time()
            if (epinemsearch %in% "greedy") {
                nemres <- nem(reddata, inference = "nem.greedy")
            } else {
                nemres <- nem(reddata, inference = "search")
            }
            nadj <- transitive.reduction(graph2adj(nemres$graph))
            time[3, i, j] <- difftime(Sys.time(), start, units = "secs")
            print(time[3, i, j])

            tp <- sum(eadj  == 1 & nadj == 1)
            tn <- sum(eadj == 0 & nadj == 0)
            fp <- sum(eadj == 0 & nadj == 1)
            fn <- sum(eadj == 1 & nadj == 0)
            sens2[3, i, j] <- tp/(tp+fn)
            spec2[3, i, j] <- tn/(tn+fp)
            print(sens2[3, i, j])
            print(spec2[3, i, j])

        }

        if ("p" %in% do) {
            print("PCalg")

            start <- Sys.time()
            pc.fit <- pc(suffStat = list(C = cor(reddata2), n = nrow(reddata2)),
                  indepTest = gaussCItest, ## indep.test: partial correlations
                  alpha=0.05, labels = colnames(reddata2), verbose = F)
            pcadj <- graph2adj(pc.fit@graph)
            time[4, i, j] <- difftime(Sys.time(), start, units = "secs")
            print(time[4, i, j])

            tp <- sum(gtnadj == 1 & pcadj == 1)
            tn <- sum(gtnadj  == 0 & pcadj == 0)
            fp <- sum(gtnadj == 0 & pcadj == 1)
            fn <- sum(gtnadj == 1 & pcadj == 0)
            sens2[4, i, j] <- tp/(tp+fn)
            spec2[4, i, j] <- tn/(tn+fp)
            print(sens2[4, i, j])
            print(spec2[4, i, j])

        }

        if ("a" %in% do) {
            print("Aracne")

            start <- Sys.time()
            ares <- build.mim(reddata2)
            ares <- aracne(ares)
            ares <- disc(ares, 0)
            ares <- ares[order(rownames(ares)), order(colnames(ares))]
            nas <- which(is.na(ares) == T)
            ares[nas] <- 0
            diag(ares) <- 0
            time[5, i, j] <- difftime(Sys.time(), start, units = "secs")
            print(time[5, i, j])

            tp <- sum(gtnadj == 1 & ares == 1)
            tn <- sum(gtnadj == 0 & ares == 0)
            fp <- sum(gtnadj == 0 & ares == 1)
            fn <- sum(gtnadj == 1 & ares == 0)
            sens2[5, i, j] <- tp/(tp+fn)
            spec2[5, i, j] <- tn/(tn+fp)
            print(sens2[5, i, j])
            print(spec2[5, i, j])

        }

    }

}
```
```{r, fig.width = 7, fig.height = 7}
data(sim)

colvec <- c(rep("orange", length(noiselvls)), rep("blue", length(noiselvls)),
            rep("darkgreen", length(noiselvls)), rep("brown", length(noiselvls)),
            rep("darkgrey", length(noiselvls)))

acc <- (sens + spec)/2

acc2 <- (sens2 + spec2)/2

m <- rbind(c(1,1), c(2,2), c(3,4))

layout(m)

timeframe <- as.data.frame(
    cbind(data.frame(epiNEM = time[1,,]),
          data.frame(BNEM = time[2,,]), data.frame(NEM = time[3,,]),
          data.frame(Cor = time[4,,]), data.frame(MI = time[5,,])))

colnames(timeframe) <- rep(noiselvls, 5)

boxplot(timeframe, col = colvec, main = "running time", ylab = "seconds")

abline(v=(1:(length(do)-1)*length(noiselvls) + 0.5), col = "black", lty = 6)

axis(1, c(3, 11, 19, 28, 36)+1, c("epiNEM", "B-NEM", "NEM", "PC Algorithm", "ARACNE"),
     tick = F, pos = -25)

accframe2 <- as.data.frame(
    cbind(data.frame(epiNEM = acc2[1,,]),
          data.frame(BNEM = acc2[2,,]), data.frame(NEM = acc2[3,,]),
          data.frame(Cor = acc2[4,,]), data.frame(MI = acc2[5,,])))

colnames(accframe2) <- rep(noiselvls, 5)

boxplot(accframe2, col = colvec, main = "accuracy of the inferred edges", ylim = c(0,1))

abline(v=(1:(length(do)-1)*length(noiselvls) + 0.5), col = "black", lty = 6)

axis(1, c(3, 11, 19, 28, 36)+1, c("epiNEM", "B-NEM", "NEM", "PC Algorithm", "ARACNE"),
     tick = F, pos = -0.2)

## logical nems:

colvec2 <- c(rep("orange", length(noiselvls)), rep("blue", length(noiselvls)))

logicsframe <- as.data.frame(cbind(data.frame(epiNEM = logics[1,,]),
                                   data.frame(BNEM = logics[2,,])))

colnames(logicsframe) <- rep(noiselvls, 2)

boxplot(logicsframe, col = colvec2, main = "accuracy of the inferred logic gate",
        ylim = c(0,1))

abline(v=length(noiselvls)+0.5, col = "black", lty = 6)

axis(1, c(3, 11, 19, 28, 36)+1, c("epiNEM", "B-NEM", "NEM", "PC Algorithm", "ARACNE"),
     tick = F, pos = -0.2)

accframe <- as.data.frame(cbind(data.frame(epiNEM = acc[1,,]),
                                data.frame(BNEM = acc[2,,])))

colnames(accframe) <- rep(noiselvls, 2)

boxplot(accframe, col = colvec2, main = "accuracy of the inferred expected data",
        ylim = c(0,1))

abline(v=length(noiselvls)+0.5, col = "black", lty = 6)

axis(1, c(3, 11, 19, 28, 36)+1, c("epiNEM", "B-NEM", "NEM", "PC Algorithm", "ARACNE"),
     tick = F, pos = -0.2)
```

## Yeast knockout screens
In this section we analyse previously published yeast knockout screens. The screens consist of gene expression data derived from double and single knockout mutants. We use epiNEM on each double mutant combined with each single mutant.\\

The results of the knockout screens have been annotated according to the following legend:

```{r, warning=FALSE, fig.width = 9, fig.height=5}
a1 <- heatmapOP(matrix(c(1,-1,1,-1,1,1, 1, 1, 1), 3, 3,
                 dimnames = list(c("A", "B", "A.B"), LETTERS[1:3])), Colv = F, Rowv = F,
          main = "OR", col = "Greys", sub = "", colorkey = NULL)
a2 <- heatmapOP(matrix(c(1,-1,1,-1,1,1, -1, -1, 1), 3, 3,
                 dimnames = list(c("A", "B", "A.B"), LETTERS[1:3])), Colv = F, Rowv = F,
          main = "AND", col = "Greys", sub = "", colorkey = NULL)
a3 <- heatmapOP(matrix(c(1,-1,1,-1,1,1, 1, -1, -1), 3, 3,
                 dimnames = list(c("A", "B", "A.B"), LETTERS[1:3])), Colv = F, Rowv = F,
          main = "B masks effect of A", col = "Greys", sub = "", colorkey = NULL)
a4 <- heatmapOP(matrix(c(1,-1,1,-1,1,1, -1, 1, -1), 3, 3,
                 dimnames = list(c("A", "B", "A.B"), LETTERS[1:3])), Colv = F, Rowv = F,
          main = "A masks effect of B", col = "Greys", sub = "", colorkey = NULL)
a5 <- heatmapOP(matrix(c(1,-1,1,-1,1,1, 1, 1, -1), 3, 3,
                 dimnames = list(c("A", "B", "A.B"), LETTERS[1:3])), Colv = F, Rowv = F,
          main = "XOR", col = "Greys", sub = "", colorkey = NULL)
a6 <- heatmapOP(matrix(c(1,-1,1,-1,1,1, -1, 1, 1), 3, 3,
                 dimnames = list(c("A", "B", "A.B"), LETTERS[1:3])), Colv = F, Rowv = F,
          main = "No epistasis", col = "Greys", sub = "", colorkey = NULL)
a7 <- heatmapOP(matrix(c(1,-1,1,-1,1,1, 1, -1, 1), 3, 3,
                 dimnames = list(c("A", "B", "A.B"), LETTERS[1:3])), Colv = F, Rowv = F,
          main = "No epistasis", col = "Greys", sub = "", colorkey = NULL)
a8 <- heatmapOP(matrix(c(1,-1,1,-1,1,1, -1, -1, -1), 3, 3,
                 dimnames = list(c("A", "B", "A.B"), LETTERS[1:3])), Colv = F, Rowv = F,
          main = "No epistasis (unconnected)", col = "Greys", sub = "", colorkey = NULL)
print(a5, position = c(0,0, .25, .5), more = TRUE)
print(a6, position = c(.25,0, .5, .5), more = TRUE)
print(a7, position = c(.5,0, .75, .5), more = TRUE)
print(a8, position = c(.75,0, 1, .5), more = TRUE)
print(a1, position = c(0,.5, .25, 1), more = TRUE)
print(a2, position = c(.25,.5, .5, 1), more = TRUE)
print(a3, position = c(.5,.5, .75, 1), more = TRUE)
print(a4, position = c(.75,.5, 1, 1))
```

### Wageningen et al., 2010

```{r}
file <-
    "http://www.holstegelab.nl/publications/sv/signaling_redundancy/downloads/DataS1.txt"

data <- read.delim(file)

dataM <- data[-(1:2), (1+(1:(324/2))*2)]

dataP <- data[-(1:2), (2+(1:(324/2))*2)]

dataM <- dataM[-1, ]

dataP <- dataP[-1, ]

dataM <- apply(dataM, c(1,2), as.numeric)

dataP <- apply(dataP, c(1,2), as.numeric)

dataBin <- dataM

sig <- 0.05

cutoff <- log2(1.7)

dataBin[which(dataP < sig & dataP > 0 & abs(dataM) >= cutoff)] <- 1

dataBin[which(dataP >= sig | dataP == 0 | abs(dataM) < cutoff)] <- 0

dataBin <- dataBin[-which(apply(dataBin, 1, max) == 0), ]

dataBinWag <- dataBin

genelist <- toupper(c('hsl1', 'cla4', 'gin4', 'swe1', 'hsl1.cla4'))

colnames(dataBin) <- gsub(".del.vs..wt", "", colnames(dataBin))

colnames(dataBin) <- gsub(".del", "", colnames(dataBin))

doubles <- colnames(dataBin)[grep("\\.", colnames(dataBin))]

doubles <- sort(doubles[-grep("vs", doubles)])

doubles.genes <- unique(unlist(strsplit(doubles, "\\.")))

singles <- colnames(dataBin)[-grep("\\.", colnames(dataBin))]

singles <- unique(sort(singles))

llmat <- logicmat <- matrix(0, length(singles), length(doubles))

rownames(llmat) <- rownames(logicmat) <- singles

colnames(llmat) <- colnames(logicmat) <- doubles

globalgenes <- which(apply(dataBin, 1, max) == 1)
```
```{r, eval=FALSE}
for (i in doubles) {
    if (which(doubles %in% i) == 8) { next() }
    print(i)
    doubles.singles <- unlist(strsplit(i, "\\."))
    egenes <- which(apply(dataBin[, which(colnames(dataBin) %in%
            c(i, doubles.singles))], 1, max) == 1)
    for (j in singles) {
        print(j)
        if (j %in% doubles.singles) { next() }

        dataTmp <- dataBin[, grep(paste(
             paste("^", c(i, j, doubles.singles), "$", sep = ""), collapse = "|"),
                                                              colnames(dataBin))]

        if (path %in% "fixed_set") {
            dataTmp <- dataTmp[egenes, ]
        }
        if (path %in% "global") {
            dataTmp <- dataTmp[globalgenes, ]
        }
        if (path %in% "") {
            dataTmp <- dataTmp[which(apply(dataTmp, 1, max) == 1), ]
        }
        
        i1 <- which(singles %in% j)
        i2 <- which(doubles %in% i)

        if (!(is.null(dim(dataTmp)))) {
            
            if (any(dataTmp[, j] != 0)) {
                
                epires <- epiNEM(dataTmp, method = "exhaustive")
                
                tmp <- epires$logics
                if ("OR" %in% tmp) {
                    if (sum(epires$origModel[, j]) != 2) {
                        tmp <- "NOEPI"
                    } else {
                        if (all(tmp %in% "OR")) {
                            tmp <- "OR"
                        } else {
                            tmp <- tmp[which(!(tmp %in% "OR"))]
                        }
                    }
                }
                
                logicmat[i1, i2] <- tmp
                llmat[i1, i2] <- epires$score
                
            } else {

                logicmat[i1, i2] <- "UNCON"
                llmat[i1, i2] <- -Inf
                
            }
            
        } else {
            
            logicmat[i1, i2] <- "UNCON"
            llmat[i1, i2] <- -Inf
            
        }
        
    }
    
}
```
```{r, fig.width = 8, fig.height = 4}
palette(c("#4444cc", "#77aa77", "#009933", "#ff0000", "#dd8811", "#aa44bb", "#999900"))

data(wageningen_res)

llmat0 <- wageningen$ll

logicmat0 <- wageningen$logic

for (i in 1:length(doubles)) {
    
    if (i %in% 8) { next() }

    logicvec <- logicmat0[, i]

    llvec <- llmat0[, i]

    logicvec <- logicvec[order(llvec, decreasing = T)]

    llvec <- llvec[order(llvec, decreasing = T)]

    parents <- unlist(strsplit(doubles[i], "\\."))

    pchvec <- numeric(length(llvec))

    pchvec[which(logicvec %in% "AND")] <- 1
    pchvec[which(logicvec %in% "OR")] <- 2
    pchvec[which(logicvec %in% "XOR")] <- 3
    pchvec[grep(paste("^", parents[1], sep = ""), logicvec)] <- 4
    pchvec[grep(paste("^", parents[2], sep = ""), logicvec)] <- 5
    pchvec[which(logicvec %in% "NOEPI")] <- 6
    pchvec[which(logicvec %in% c("NOINFO", "NOINF"))] <- 7
    
    logicvec <- logicvec[-which(logicvec %in% "0")]
    pchvec <- pchvec[-which(pchvec == 0)]
    llvec <- llvec[-which(llvec == 0)]

    colvec <- pchvec
    
    if (all(is.infinite(llvec) == T)) {

        llvec[1:length(llvec)] <- -1000

        margin <- 100

        donames <- 30

    } else {
        
        llvec[which(is.infinite(llvec) == T)] <- NA
        
        ## llvec[which(is.infinite(llvec) == T)] <- min(llvec) - 100
        
        margin <- abs(max(llvec[1:30], na.rm = T) - min(llvec[1:30], na.rm = T))

        offset <- 0.075

        if (margin == 0) { margin <- 10; offset <- 0.0375 }

        donames <- 30 - sum(is.na(llvec[1:30]) == T)

        if (any(is.na(llvec[1:30]) == T)) { margin2 <- margin*2
        } else { margin2 <- margin }
        
        llvec[which(is.na(llvec) == T)] <- min(llvec, na.rm = T) - margin
        
        margin <- margin2

    }

    if (all(llvec[-(1:30)] - min(llvec[-(1:30)]) == 0)) {

        p2max <- max(llvec[-(1:30)]) + margin

    } else {

        p2max <- max(llvec[-(1:30)])
        
    }
    
    mark <- ""
    thetop <- sum(!(logicvec %in% c("NOINFO", "NOINF")))
    legendx <- length(llvec[1:thetop])
    p2max <- max(llvec[1:thetop])
    if (p2max == min(llvec[1:thetop])) {
        p2max <- p2max+margin*0.2
    }
    legendtext <- c("AND", "OR", "XOR", paste(parents[1]," masks ", parents[2], sep = ""),
                    paste(parents[2], " masks ", parents[1], sep = ""), "no epistasis")
    if (thetop == 0) { next() }
    plot = plot(llvec[1:thetop], pch = pchvec[1:thetop], col = colvec[1:thetop],
                ylab = "likelihood", xlab = "ranked single knockouts",
                ylim = c(min(llvec[1:thetop]), max(llvec[1:thetop])+margin*0.2),
                xlim = c(1, thetop+(thetop/100)),
                main = paste(unlist(strsplit(doubles[i], "\\.")), collapse = " and "))
    text = text((1:thetop)+(thetop/100), llvec[1:thetop]+(margin*offset),
                labels = names(llvec)[1:thetop], cex = 0.6, srt = 45, pos = 3,
                offset = 0)
    mtext = mtext(mark, side = 3, line = 1, outer = F, cex = 4, adj = 0)
    legend = legend(legendx, p2max,
                    legend = legendtext,
                    col = 1:6, pch = 1:6, xjust = 1, yjust = 1, cex = 0.7)
    
}

distmat <- wageningen$logic

distmat[which(distmat %in% "AND")] <- 1
distmat[which(distmat %in% "OR")] <- 2
distmat[which(distmat %in% "XOR")] <- 3
distmat[which(distmat %in% "NOEPI")] <- 6
distmat[which(distmat %in% c("NOINFO", "NOINF"))] <- 7

for (i in 1:ncol(distmat)) {

    genes <- unlist(strsplit(colnames(distmat)[i], "\\."))
    
    distmat[which(distmat[, i] %in%
                  paste(genes[1], " masks the effect of ", genes[2], sep = "")), i] <- 4

    
    distmat[which(distmat[, i] %in%
                  paste(genes[2], " masks the effect of ", genes[1], sep = "")), i] <- 5

}

distmat <- apply(distmat, c(1,2), as.numeric)

for (i in 1:ncol(distmat)) {
    distmat[, i] <- rev(sort(distmat[, i]))
}

distmat <- distmat[-which(apply(distmat, 1, sum) == 0), ]

distmat <- distmat[, -which(apply(distmat, 2, max) == 0 | apply(distmat, 2, min) == 7)]

y <- distmat

distmat <- distmat[, order(apply(distmat, 2, function(x) { return(sum(x == 1)) }))]

y[which(y == 5)] <- 4

rownames(distmat) <- NULL

labeltext <- c("", "no information\n\n\n", "no epistasis\n\n\n","masking (NOT B)\n\n\n",
              "masking (NOT A)\n\n\n", "XOR\n\n\n", "OR\n\n\n", "AND\n\n\n")

heatmapOP(distmat, Colv = F, Rowv = F, main = "logic gate distribution",
          sub = "", col = "Paired", breaks = seq(0.5,7.5, length.out = 8),
          cexRow = 0, cexCol = 0.4, aspect = "fill",
          colorkey = list(space = "right",
                          labels = rev(labeltext), width = 1,
                          at = seq(1.5,7.5, length.out = 8)),
          xlab = "double knock-outs",
          ylab = "modulators\n(different order for each pair)",
          xrot = 45, bordercol = "transparent")
```

### Sameith et al., 2015

```{r}
file <- paste("http://www.holstegelab.nl/publications/GSTF_geneticinteractions/",
              "downloads/del_mutants_limma.txt", sep = "")

data <- read.delim(file)

data <- apply(data, c(1,2), as.character)

dataM <- data[-1, which(data[1, ] %in% "M")]

dataM <- apply(dataM, c(1,2), as.numeric)

dataP <- data[-1, which(data[1, ] %in% "p.value")]

dataP <- apply(dataP, c(1,2), as.numeric)

dataBin <- dataM

sig <- 0.01

cutoff <- log2(1.5)

dataBin[which(dataP < sig & dataP > 0 & abs(dataM) >= cutoff)] <- 1

dataBin[which(dataP >= sig | dataP == 0 | abs(dataM) < cutoff)] <- 0

dataBin <- dataBin[-which(apply(dataBin, 1, max) == 0), ]

colnames(dataBin) <- gsub("\\.\\.\\.", "\\.", colnames(dataBin))

## big screen:

doubles <- colnames(dataBin)[grep("\\.", colnames(dataBin))]

doubles.genes <- unique(unlist(strsplit(doubles, "\\.")))

singles <- colnames(dataBin)[-grep("\\.", colnames(dataBin))]

singles <- unique(sort(singles))

llmat <- logicmat <- matrix(0, length(singles), length(doubles))

rownames(llmat) <- rownames(logicmat) <- singles

colnames(llmat) <- colnames(logicmat) <- doubles

globalgenes <- which(apply(dataBin, 1, max) == 1)
```
```{r, eval=FALSE}
for (i in doubles[set]) {
    print(i)
    doubles.singles <- unlist(strsplit(i, "\\."))
    egenes <- which(apply(dataBin[,
                which(colnames(dataBin) %in% c(i, doubles.singles))], 1, max) == 1)
    for (j in singles) {
        print(j)
        if (j %in% doubles.singles) { next() }

        dataTmp <- dataBin[, grep(paste(paste("^", c(i, j, doubles.singles), "$", sep
                    = ""), collapse = "|"), colnames(dataBin))]

        if (path %in% "fixed_set") {
            dataTmp <- dataTmp[egenes, ]
        }
        if (path %in% "global") {
            dataTmp <- dataTmp[globalgenes, ]
        }
        if (path %in% "") {
            dataTmp <- dataTmp[which(apply(dataTmp, 1, max) == 1), ]
        }
        
        i1 <- which(singles %in% j)
        i2 <- which(doubles %in% i)

        if (!(is.null(dim(dataTmp)))) {
            
            if (any(dataTmp[, j] != 0)) {
                
                epires <- epiNEM(dataTmp, method = "exhaustive")
                
                tmp <- epires$logics
                if ("OR" %in% tmp) {
                    if (sum(epires$origModel[, j]) != 2) {
                        tmp <- "NOEPI"
                    } else {
                        if (all(tmp %in% "OR")) {
                            tmp <- "OR"
                        } else {
                            tmp <- tmp[which(!(tmp %in% "OR"))]
                        }
                    }
                }
                
                logicmat[i1, i2] <- tmp
                llmat[i1, i2] <- epires$score
                
            } else {

                logicmat[i1, i2] <- "UNCON"
                llmat[i1, i2] <- -Inf
                
            }
            
        } else {
            
            logicmat[i1, i2] <- "UNCON"
            llmat[i1, i2] <- -Inf
            
        }
        
    }

}
```
```{r, fig.width = 8, fig.height = 4}
data(sameith_res)

llmat0 <- sameith$ll

logicmat0 <- sameith$logic

for (i in 1:length(doubles)) {
  
  logicvec <- logicmat0[, i]

  llvec <- llmat0[, i]

  logicvec <- logicvec[order(llvec, decreasing = T)]

  llvec <- llvec[order(llvec, decreasing = T)]

  parents <- unlist(strsplit(doubles[i], "\\."))

  pchvec <- numeric(length(llvec))

    pchvec[which(logicvec %in% "AND")] <- 1
    pchvec[which(logicvec %in% "OR")] <- 2
    pchvec[which(logicvec %in% "XOR")] <- 3
    pchvec[grep(paste("^", parents[1], sep = ""), logicvec)] <- 4
    pchvec[grep(paste("^", parents[2], sep = ""), logicvec)] <- 5
    pchvec[which(logicvec %in% "NOEPI")] <- 6
    pchvec[which(logicvec %in% c("NOINFO", "NOINF"))] <- 7

    logicvec <- logicvec[-which(logicvec %in% "0")]
    pchvec <- pchvec[-which(pchvec == 0)]
    llvec <- llvec[-which(llvec == 0)]

    colvec <- pchvec

    if (all(is.infinite(llvec) == T)) {

        llvec[1:length(llvec)] <- -1000

        margin <- 100

        donames <- 30

    } else {
        
        llvec[which(is.infinite(llvec) == T)] <- NA
        
        margin <- abs(max(llvec[1:30], na.rm = T) - min(llvec[1:30], na.rm = T))

        if (margin == 0) { margin <- 10 }

        donames <- 30 - sum(is.na(llvec[1:30]) == T)

        if (any(is.na(llvec[1:30]) == T)) { margin2 <- margin*2
        } else { margin2 <- margin }
        
        llvec[which(is.na(llvec) == T)] <- min(llvec, na.rm = T) - margin
        
        margin <- margin2

    }

    if (all(llvec[-(1:30)] - min(llvec[-(1:30)]) == 0)) {

        p2max <- max(llvec[-(1:30)]) + margin

    } else {

        p2max <- max(llvec[-(1:30)])
        
    }
    
    labeltext <- c("AND", "OR", "XOR", paste(parents[1], " masks ", parents[2], sep = ""),
                   paste(parents[2], " masks ", parents[1], sep = ""), "no epistasis")

    mark <- ""
    pointx <- 10000
    thetop <- sum(!(logicvec %in% c("NOINFO", "NOINF")))
    legendx <- length(llvec[1:thetop])
    p2max <- max(llvec[1:thetop])
    if (p2max == min(llvec[1:thetop])) {
        p2max <- p2max+margin*0.2
    }
    if (thetop == 0) { next() }
    plot = plot(llvec[1:thetop], pch = pchvec[1:thetop], col = colvec[1:thetop],
                ylab = "likelihood", xlab = "ranked single knockouts",
                ylim = c(min(llvec[1:thetop]), max(llvec[1:thetop])+margin*0.2),
                xlim = c(1, thetop+(thetop/100)),
                main = paste(tolower(unlist(strsplit(doubles[i], "\\."))),
                             collapse = " and "))
    text = text((1:thetop)+(thetop/100), llvec[1:thetop]+(margin*offset),
                labels = tolower(names(llvec)[1:thetop]), cex = 0.6, srt = 45, pos = 3,
                offset = 0)
    mtext = mtext(mark, side = 3, line = 1, outer = F, cex = 4, adj = 0)
    legend = legend(legendx, p2max,
                    legend = labeltext, col = 1:6, pch = 1:6, xjust = 1, yjust = 1,
                    cex = 0.7)  
    
}

distmat <- sameith$logic

distmat[which(distmat %in% "AND")] <- 1
distmat[which(distmat %in% "OR")] <- 2
distmat[which(distmat %in% "XOR")] <- 3
distmat[which(distmat %in% "NOEPI")] <- 6
distmat[which(distmat %in% c("NOINFO", "NOINF"))] <- 7

for (i in 1:ncol(distmat)) {

    genes <- unlist(strsplit(colnames(distmat)[i], "\\."))
    
    distmat[which(distmat[, i] %in% paste(genes[1], " masks the effect of ",
                                          genes[2], sep = "")), i] <- 4

    
    distmat[which(distmat[, i] %in% paste(genes[2], " masks the effect of ",
                                          genes[1], sep = "")), i] <- 5

}

distmat <- apply(distmat, c(1,2), as.numeric)

for (i in 1:ncol(distmat)) {
    distmat[, i] <- rev(sort(distmat[, i]))
}

distmat <- distmat[-which(apply(distmat, 1, sum) == 0), ]

library(bnem)

y <- distmat

distmat <- distmat[, order(apply(distmat, 2, function(x) { return(sum(x == 1)) }))]

y[which(y == 5)] <- 4

rownames(distmat) <- NULL

labeltext <- c("", "no information\n\n\n", "no epistasis\n\n\n",
               "masking (NOT B)\n\n\n", "masking (NOT A)\n\n\n",
               "XOR\n\n\n", "OR\n\n\n", "AND\n\n\n")

heatmapOP(distmat, Colv = F, Rowv = F, main = "logic gate distribution", sub = "",
          col = "Paired", breaks = seq(0.5,7.5, length.out = 8), cexRow = 0,
          cexCol = 0.4, aspect = "fill",
          colorkey = list(space = "right", labels = rev(labeltext), width = 1,
                          at = seq(1.5,7.5, length.out = 8)),
          xlab = "double knock-outs",
          ylab = "modulators\n(different order for each pair)",
          xrot = 45, bordercol = "transparent")
```

Now we plot the densities of the string-db interaction scores of our identified modulators and a random draw.

```{r, fig.width = 8, fig.height = 4}
par(mfrow=c(1,2))

library(STRINGdb)

get_STRING_species(version="10", species_name=NULL)[26, ] # 4932

string_db <- STRINGdb$new( version="10", species=4932, score_threshold=0,
                          input_directory="~/")

llmat <- wageningen$ll

logicmat <- wageningen$logic

string.scores <- list()

string.names <- character()
```
```{r, eval=FALSE}
for (i in 1:ncol(llmat)) {

    if (sum(!(llmat[, i] %in% c(0,-Inf))) > 0) { 
        top30 <- llmat[, i]
        top30[which(top30 == 0)] <- -Inf
        top30 <- top30[which(!(llmat[, i] %in% c(0,-Inf)))]
        top30 <- top30[order(top30,decreasing = T)[1:min(30, sum(!(llmat[, i]
            %in% c(0,-Inf))))]]

        doubles <- unlist(strsplit(colnames(llmat)[i], "\\."))

        for (j in names(top30)) {
            tmp <- string_db$get_interactions(string_db$mp(c(doubles[1], j)))
            string.scores <- c(string.scores, tmp$combined_score)
            string.names <- c(string.names, paste(sort(c(doubles[1], j)), collapse = "_"))
            tmp <- string_db$get_interactions(string_db$mp(c(doubles[2], j)))
            string.scores <- c(string.scores, tmp$combined_score)
            string.names <- c(string.names, paste(sort(c(doubles[2], j)), collapse = "_"))
        }
        
    } else {
        next()
    }

}
```
```{r, fig.width = 8, fig.height = 5}
data(wageningen_string)

tmp <- string_db$get_interactions(string_db$mp(unique(unlist(strsplit(colnames(dataBinWag)
                                                                     , "\\.")))))

stsc <- unlist(string.scores)

denspval <- wilcox.test(stsc, unlist(tmp$combined_score), alternative = "greater")$p.value

for (i in 100:1) {

    if (denspval < 10^(-i)) {

        denspval <- paste("< ", 10^(-i), sep = "")

    } 

}

plot(density(stsc), col = "#00000000",
     ylim = c(0, max(c(max(density(stsc)$y),max(density(unlist(tmp$combined_score))$y)))),
     main = paste("Mann-Whitney test p-value ", denspval, sep = ""), xlab = "",
     cex.main = 1.5)
polygon(density(stsc), col = "#ff000066")

lines(density(unlist(tmp$combined_score)), col = "#00000000")
polygon(density(unlist(tmp$combined_score)), col = "#00ffff66")

llmat <- sameith$ll

logicmat <- sameith$logic

string.scores2 <- list()

string.names2 <- character()
```
```{r, eval=FALSE}
for (i in 1:ncol(llmat)) {

    if (sum(!(llmat[, i] %in% c(0,-Inf))) > 0) { 
        top30 <- llmat[, i]
        top30[which(top30 == 0)] <- -Inf
        top30 <- top30[which(!(llmat[, i] %in% c(0,-Inf)))]
        top30 <- top30[order(top30, decreasing = T)[1:min(30, sum(!(llmat[, i]
            %in% c(0,-Inf))))]]

        doubles <- unlist(strsplit(colnames(llmat)[i], "\\."))

        for (j in names(top30)) {
            tmp <- string_db$get_interactions(string_db$mp(c(doubles[1], j)))
            string.scores2 <- c(string.scores2, tmp$combined_score)
            string.names2 <- c(string.names2, paste(sort(c(doubles[1], j)), collapse = "_"))
            tmp <- string_db$get_interactions(string_db$mp(c(doubles[2], j)))
            string.scores2 <- c(string.scores2, tmp$combined_score)
            string.names2 <- c(string.names2, paste(sort(c(doubles[2], j)), collapse = "_"))
        }
        
    } else {
        next()
    }

}
```
```{r, fig.width = 8, fig.height = 5}
data(sameith_string)

tmp <- string_db$get_interactions(string_db$mp(unique(unlist(strsplit(colnames(dataBin)
                                                                    , "\\.")))))

stsc <- unlist(string.scores2)

denspval <- wilcox.test(stsc, unlist(tmp$combined_score), alternative = "greater")$p.value

for (i in 100:1) {

    if (denspval < 10^(-i)) {

        denspval <- paste("< ", 10^(-i), sep = "")

    } 

}

plot(density(stsc), col = "#00000000",
     ylim = c(0, max(c(max(density(stsc)$y), max(density(unlist(tmp$combined_score))$y)))),
     main = paste("Mann-Whitney test p-value ", denspval, sep = ""), xlab = "",
     cex.main = 1.5)
polygon(density(stsc), col = "#ff000066")

lines(density(unlist(tmp$combined_score)), col = "#00000000")
polygon(density(unlist(tmp$combined_score)), col = "#00ffff66")
```

```{r}
sessionInfo()
```

*__Reference:__*

Martin Pirkl, Madeline Diekmann, Marlies van der Wees, Niko Beerenwinkel, Holger Fröhlich, Florian Markowetz. Inferring Modulators of Genetic Interactions with Epistatic Nested Effects Models. submitted.
